---
title: "PsyEmbedR Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PsyEmbedR Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`PsyEmbedR` supports a local semantic auditing workflow for psychometric
item banks. The package combines R and Python through `reticulate` and
uses `sentence-transformers` to produce sentence embeddings without
using external AI APIs.

## 1. Load the package

```{r}
library(PsyEmbedR)
```

## 2. Prepare a Python environment

The helper below creates a virtual environment if needed, installs the
required Python libraries, activates the environment, and verifies that
`sentence_transformers` is available.

```{r, eval = FALSE}
ensure_python_deps()
```

The virtual environment contains:

- `sentence-transformers`
- `torch`
- `numpy`
- `scikit-learn`

If you are configuring the package on a new machine and the default
virtualenv root does not exist yet, you can point `reticulate` to a
writable directory before loading the package:

```{r, eval = FALSE}
Sys.setenv(RETICULATE_VIRTUALENV_ROOT = "C:/Users/your-user/Documents/.virtualenvs")
library(PsyEmbedR)
ensure_python_deps()
```

## 3. Define items and theoretical definitions

```{r}
items <- c(
  "I adapt my plans when the context changes unexpectedly.",
  "I detect relevant environmental changes very quickly.",
  "I recover my concentration after a difficult event.",
  "I modify my behavior when demands increase."
)

definitions <- c(
  adaptive = "Flexible behavior change in response to situational demands.",
  ecological = "Detection of relevant cues in the surrounding environment.",
  resilient = "Recovery of effective functioning after challenge or stress."
)
```

## 4. Detect language and resolve the model

`PsyEmbedR` includes a lightweight local heuristic for short
psychometric texts:

```{r}
detect_text_language(c(items, unname(definitions)))
```

The main semantic functions default to `model_name = "auto"`. In that
mode, English-only text is routed to
`sentence-transformers/all-mpnet-base-v2`, while Spanish, mixed, or
uncertain text is routed to
`sentence-transformers/paraphrase-multilingual-mpnet-base-v2`.

## 5. Generate embeddings

```{r, eval = FALSE}
embeddings <- embed_items(items)
dim(embeddings)
```

To use a different model, pass `model_name` explicitly. For fully
offline work, provide a local model path or a model that has already
been cached on the machine.

## 6. Compute item-to-item semantic similarity

```{r, eval = FALSE}
item_similarity <- cosine_matrix(embeddings)
item_similarity
```

This matrix is useful for identifying clusters of very similar items
before empirical fieldwork.

## 7. Align items to theoretical definitions

```{r, eval = FALSE}
alignment <- align_to_definition(
  items = items,
  definitions = definitions
)

alignment
```

The returned tibble is in long format and includes:

- the item identifier,
- the item text,
- each candidate definition,
- the cosine similarity for that pairing,
- the best semantic match for each item, and
- the best similarity score for each item.

## 8. Infer the most likely dimension for each item

```{r, eval = FALSE}
infer_dimensions(
  items = items,
  definitions = definitions
)
```

This is a convenience wrapper over `align_to_definition()` that returns
the best semantic classification for each item.

By default, if the best cosine similarity is below `0.20`, the item is
flagged as lacking a clear semantic alignment and receives the label
`"no clear alignment"`. You can adjust that threshold with
`min_similarity`.

The output distinguishes between:

- `nearest_definition`: the closest theoretical definition before
  applying the threshold.
- `predicted_dimension`: the final threshold-aware preliminary
  assignment shown in the report.

## 9. Visualize the semantic map in two dimensions

```{r, eval = FALSE}
plot_semantic_map(items)
```

The default projection uses classical multidimensional scaling over
cosine distances, so the layout is linked directly to the item-to-item
similarity structure. If needed, you can also request `method = "pca"`.
Label placement defaults to `label_method = "auto"`, which uses
repelled labels when `ggrepel` is available and otherwise falls back to
plain text labels.

Coloring is independent of the projection method. For example:

```{r, eval = FALSE}
plot_semantic_map(
  items = items,
  definitions = definitions,
  color_by = "inferred"
)

plot_semantic_map(
  items = items,
  expected_dimension = c("adaptive", "ecological", "resilient", "adaptive"),
  method = "pca",
  color_by = "expected"
)
```

## 10. Flag potential redundancy

```{r, eval = FALSE}
flag_redundancy(items, threshold = 0.85)
```

Pairs at or above the threshold should be reviewed for semantic overlap
and possible item pruning.

## 11. Summarize semantic coverage by dimension

```{r, eval = FALSE}
coverage <- semantic_coverage_report(
  items = items,
  definitions = definitions,
  expected_dimension = c("adaptive", "ecological", "resilient", "adaptive")
)

coverage$summary_by_dimension
coverage$classification_summary
coverage$misaligned_items
```

This report helps determine whether a bank appears semantically balanced
across dimensions and whether item assignments are consistent with their
intended theoretical labels.

## 12. Example scripts in Spanish and English

The package includes two complete step-by-step scripts:

- `inst/examples/workflow_spanish.R`
- `inst/examples/workflow_english.R`

## 13. Recommended interpretation

`PsyEmbedR` is intended for preliminary semantic review. It should be
used to complement, not replace, substantive theoretical review and
empirical psychometric validation.
